Equivalence Hypergraphs: E-Graphs for Monoidal Theories

### 2023-12-20 sections/soundness_and_completeness.tex
Soundness and fullness for PROP(\Sigma)+

\update{Below I kind of use $\llbracket - \rrbracket$ to define itself when using $\llbracket \mathcal{E} \rrbracket$.}
In this section, we will build a correspondence between terms in a semilattice enriched $\textsf{PROP}^{+}(\Sigma)$ and morphisms in $\WellTypedMdaEcospans$.
Similarly to $\textsf{PROP}^{+}(\Sigma)$ where terms are defined up to semilattice enrichment and SMC equations we will say that two morphisms $G$ and $F$ in $\WellTypedMdaEcospans$ are equal when $G \Rrightarrow{}_{\mathcal{S}} F$. 
As SMC equations are absorbed by e-hypergraph representation we only quotient by $\mathcal{S}$. 
We are now ready to build a semilattice-enriched SMC functor

\[
[\cdot] : \textsf{PROP}^{+}(\Sigma) \to \sfrac{\WellTypedMdaEcospans}{\mathcal{S}}
\]
where $\sfrac{\WellTypedMdaEcospans}{\mathcal{S}}$ is $\WellTypedMdaEcospans$ quotiented by structural rewrite rules which makes this category into a semilattice-enrhiched one.
By freeness, it suffices to define how the functor acts on generators \update{Identities and generators from the signature?}.
% As $\textsf{PROP}(\Sigma)$ and $\textsf{PROP}^{+}(\Sigma)$ essentially have the same set of generators we can reuse the functor $\llbracket \cdot \rrbracket: \textsf{PROP}(\Sigma) \to \MdaCospans$ from~\cite{Frobenius2} and then follow the inclusion $\mathcal{I}$ defined in~\ref{functor:I}.
Everything else follows by defining

\begin{enumerate}
    \item $[ f;g ] = [ f ] ; [ g ]$  for $f,g \in \Sigma$;
    \item $[ f \otimes g ] = [ f ] \otimes [ g ]$  for $f,g \in \Sigma$;
    \item $[ f + g ] = [ f ] + [ g ]$  for $f,g \in \Sigma$;
    \item \update{Symmetry?}
\end{enumerate}

this also makes the functor into a symmetric monoidal semilattice-enriched SMC one on the nose.
% \update{This essentially means that $[\cdot] = \llbracket \cdot \rrbracket ; \mathcal{I}$ plus three equations above.}

\update{Anything about soundness? I.e. if $f = g$ then $\llbracket f \rrbracket = \llbracket g \rrbracket$? Because $f = g$ is modulo semilattice-enriched SMC equations we have to show that $\llbracket f \rrbracket \Rrightarrow \llbracket g \rrbracket$. Doesn't it require the soundness of rewriting?
}

\begin{theorem}
    If $f = g$ in $PROP^{+}(\Sigma)$ then $[f] \Rrightarrow_{\mathcal{S}} [g]$ in $\WellTypedMdaEcospans$.
    \begin{proof}
        \update{
            By induction on the number of applied equations to turn $f$ into $g$?
            E.g. $f = g$ in one step essentially means that $f = l$ for some $l$ in~\ref{fig:string-equations} and $g = r$ (or the other way around).
            Then we need to show that $[f] \Rrightarrow_{\langle l, r \rangle \in \mathcal{S}} [g']$ and $[g'] = [g]$.
            }
        \question{Can we use normal form here as well? $f = g$ means there exists $h$ such that $f =_{SMC^{+}} h$ and $h =_{SMC^{+}}$ and $h$ is of the form $\phi_1 + \ldots + \phi_n$?
                  Then we need to show that $[f] \Rrightarrow_{\mathcal{S}} [h]$ and $[h] \Rrightarrow_{\mathcal{S}} [g]$.
        }
    \end{proof}
\end{theorem}

\begin{corollary}
    $[\cdot] : PROP^{+}(\Sigma) \to \sfrac{\WellTypedMdaEcospans}{\mathcal{S}}$ is sound in a sense that
    if $f = g$ in $PROP^{+}(\Sigma)$ then $[f] = [g]$ in $\sfrac{\WellTypedMdaEcospans}{\mathcal{S}}$
\end{corollary}

\begin{proposition}
    Functor $\llbracket \cdot \rrbracket; \mathcal{I}$ is faithful.
    \begin{proof}
        Consider the commutative diagram below \update{Is it obvious that the diagram commutes?}.

        % https://q.uiver.app/#q=WzAsNCxbMCwwLCJcXHRleHR7UFJPUH0oXFxTaWdtYSkiXSxbMiwwLCJcXHRleHR7UFJPUH1eeyt9KFxcU2lnbWEpIl0sWzAsMiwiXFxNZGFDb3NwYW5zIl0sWzIsMiwiXFxzZnJhY3tcXFdlbGxUeXBlZE1kYUVjb3NwYW5zfXtcXG1hdGhjYWx7U319Il0sWzAsMSwiXFxtYXRoY2Fse0p9Il0sWzAsMiwiXFxsbGJyYWNrZXQgXFxjZG90IFxccnJicmFja2V0IiwyXSxbMSwzLCJbXFxjZG90XSJdLFsyLDMsIlxcbWF0aGNhbHtJfSIsMl1d
        \[
        \begin{tikzcd}
            {\text{PROP}(\Sigma)} && {\text{PROP}^{+}(\Sigma)} \\
            \\
            \MdaCospans && {\sfrac{\WellTypedMdaEcospans}{\mathcal{S}}}
            \arrow["{\mathcal{J}}", from=1-1, to=1-3]
            \arrow["{\llbracket \cdot \rrbracket}"', from=1-1, to=3-1]
            \arrow["{[\cdot]}", from=1-3, to=3-3]
            \arrow["{\mathcal{I}}"', from=3-1, to=3-3]
        \end{tikzcd}
        \]
    \end{proof}

@@

\begin{theorem}[$\llbracket - \rrbracket$ is faithful]
\begin{proof}
Essentially we need to show that if $\llbracket f \rrbracket = \llbracket g \rrbracket$ then $f = g$ which is equivalent to if $f \not = g$ then $\llbracket f \rrbracket \not = \llbracket g \rrbracket$.
First, we note that we have a \question{weak (modulo commutativity)} normal form for any term \update{string diagram?} in $PROP^{+}(\Sigma)$ which is achieved by applying distributivity for tensor and composition as well as unpacking equations, that is, every term is representable as in Figure~\ref{fig:normal_form_string} where $\phi_1,\;\ldots\;,\phi_n$ are pure SMC terms.
\update{It is not actaully a `normal' form because we can still apply SMC equations, so it should be normal modulo SMC and commutativity}.

@@

@@

We will next consider functors
\[
    \textsf{PROP}(\Sigma, \mathcal{E}) \xrightarrow{\mathcal{J}} \textsf{PROP}^{+}(\Sigma, \mathcal{E}) \xrightarrow{\llangle - \rrangle} \sfrac{\WellTypedMdaEcospans}{\mathcal{S},\mathcal{E}}
\]
where $\mathcal{E}$ is a set of equations of the corresponding SMT.
\begin{theorem}[$\llangle - \rrangle$ is a functor]
\end{theorem}

\begin{theorem}[$\llangle - \rrangle$ is faithful]
\end{theorem}

\update{Or we can define rewriting for string diagrams in $\textsf{PROP}^{+}(\Sigma)$ based on rewriting in $\textsf{PROP}(\Sigma)$ and then say that $f \xrightarrow{}_{\mathcal{R}} g$ iff $[f] [g]$}

@@

\begin{theorem}
    $a \Rightarrow_{\langle l, r \rangle}^{+} b$ iff $[a] \Rrightarrow_{\langle [l],[r] \rangle} [b]$.
\end{theorem}

>>>>

In this section, we will build a correspondence between terms in a semilattice enriched $\textsf{PROP}^{+}(\Sigma)$ and morphisms in $\WellTypedMdaEcospans$.
% Similarly to $\textsf{PROP}^{+}(\Sigma)$ where terms are defined up to semilattice enrichment and SMC equations we will say that two morphisms $G$ and $F$ in $\WellTypedMdaEcospans$ are equal when $G \Rrightarrow{}_{\mathcal{S}} F$. 
% As SMC equations are absorbed by e-hypergraph representation we only quotient by $\mathcal{S}$. 
We will build a semilattice-enriched SMC functor

\[
[-] : \textsf{PROP}^{+}(\Sigma) \to \sfrac{\WellTypedMdaEcospans}{\mathcal{S}}
\]
where $\sfrac{\WellTypedMdaEcospans}{\mathcal{S}}$ is $\WellTypedMdaEcospans$ quotiented by structural rewrite rules $\mathcal{S} = \{\langle [l], [r] \rangle, \langle [r],[l] \rangle\}$ for all $l = r$ in~\ref{fig:string-equations}.
By freeness, it suffices to define how the functor acts on generators.
For every $f : n \to m \in \Sigma$, $[f]$ is defined as below.

\[
\scalebox{0.75}{
\tikzfig{soundness_and_completeness/generator_translation}
}
\]

\begin{theorem}
    If $f = g$ in $PROP^{+}(\Sigma)$ then $[f] \Rrightarrow_{\mathcal{S}} [g]$ in $\WellTypedMdaEcospans$.
    \begin{proof}
        $f = g$ means that $f$ as a term can be turned into $g$ by applying SMC equations and the equations from~\ref{fig:string-equations}.
        Since $f_1 =_{\text{SMC}} f_2$ means that $[f_1] = [f_2]$ we will only consider the transformations induced by the equations from~\ref{fig:string-equations}.
        We will also note that every subterm of $f$ corresponds to a convex down-closed sub e-hypergraph of $[f]$.
        Then, proving the above statement boils down to showing that for every rewrite rule in $\mathcal{S}$ the corresponding pushout complement and pushout exist.
        Suppose, $f =_{r} g$ where $r$ is the top left equation from~\ref{fig:string-equations}.
        $[r]$ is then given below.
        \[
            \scalebox{0.45}{
                \tikzfig{combinatorial_semantics/structural_rule_pushout}
            }
        \]
        After we identify a convex down-closed image of the left-hand side of this rule within a graph $[f]$ we remove the image by keeping only the outermost nodes.
        Note that because the image of all $a$'s and $b$'s are outermost nodes, we are allowed to pick any monomorphism for our pushout complement.
        And for the same reason the pushout for the right square exists.
        Then we simply glue the right-hand side of the rule into the hole.
        Similar argument applies to other rules as well.
        The result for any number of applied equations follows by induction.
    \end{proof}
\end{theorem}

\begin{corollary}
    $\sfrac{\WellTypedMdaEcospans}{\mathcal{S}}$ is a semilattice-enriched SMC category and $[-]$ is a semilattice-enriched SMC functor.
\end{corollary}

\begin{proposition}
    Functor $\llbracket \cdot \rrbracket; \mathcal{I}$ is faithful.
    \begin{proof}
        Consider the commutative diagram below.

        % https://q.uiver.app/#q=WzAsNCxbMCwwLCJcXHRleHR7UFJPUH0oXFxTaWdtYSkiXSxbMiwwLCJcXHRleHR7UFJPUH1eeyt9KFxcU2lnbWEpIl0sWzAsMiwiXFxNZGFDb3NwYW5zIl0sWzIsMiwiXFxzZnJhY3tcXFdlbGxUeXBlZE1kYUVjb3NwYW5zfXtcXG1hdGhjYWx7U319Il0sWzAsMSwiXFxtYXRoY2Fse0p9Il0sWzAsMiwiXFxsbGJyYWNrZXQgXFxjZG90IFxccnJicmFja2V0IiwyXSxbMSwzLCJbXFxjZG90XSJdLFsyLDMsIlxcbWF0aGNhbHtJfSIsMl1d
        \[
        \begin{tikzcd}
            {\text{PROP}(\Sigma)} && {\text{PROP}^{+}(\Sigma)} \\
            \\
            \MdaCospans && {\sfrac{\WellTypedMdaEcospans}{\mathcal{S}}}
            \arrow["{\mathcal{J}}", from=1-1, to=1-3]
            \arrow["{\llbracket \cdot \rrbracket}"', from=1-1, to=3-1]
            \arrow["{[-]}", from=1-3, to=3-3]
            \arrow["{\mathcal{I}}"', from=3-1, to=3-3]
        \end{tikzcd}
        \]
    \end{proof}

@@

\begin{theorem}
$[-]$ is faithful.
\begin{proof}
Essentially we need to show that if $[f] = [g]$ then $f = g$ which is equivalent to if $f \not = g$ then $[f] \not = [g]$.
First, we note that we have a \question{weak (modulo commutativity)} normal form for any term \update{string diagram?} in $PROP^{+}(\Sigma)$ which is achieved by applying distributivity for tensor and composition as well as unpacking equations, that is, every term is representable as in Figure~\ref{fig:normal_form_string} where $\phi_1,\;\ldots\;,\phi_n$ are pure SMC terms.

@@

\begin{theorem}
$[-]$ is full.
\begin{proof}
As we quotient by $\mathcal{S}$ instead of arbitrary cospans we can only consider their normal forms $A_1 + \ldots A_n$ where all of $A_i$ do not contain any hierarchical edges.
Because each of $A_i$ is a plain hypergraph, they have a pre-image in $\textsf{PROP}(\Sigma)$, i.e. $A_i = \llbracket \cdot \rrbracket;\mathcal{I}(a_i)$.
And then $[\mathcal{J}(a_1) + \ldots + \mathcal{J}(a_n)] = [\mathcal{J}(a_1)] + \ldots + [\mathcal{J}(a_n)] = A_1 + \ldots + A_n$ and hence $\mathcal{J}(a_1) + \ldots + \mathcal{J}(a_n)$ is the pre-image of $A_1 + \ldots + A_n$.
\end{proof}
\end{theorem}

@@

% We will next consider functors
% \[
%     \textsf{PROP}(\Sigma, \mathcal{E}) \xrightarrow{\mathcal{J}} \textsf{PROP}^{+}(\Sigma, \mathcal{E}) \xrightarrow{\llangle - \rrangle} \sfrac{\WellTypedMdaEcospans}{\mathcal{S},\mathcal{E}}
% \]
% where $\mathcal{E}$ is a set of equations of the corresponding SMT.
% \begin{theorem}[$\llangle - \rrangle$ is a functor]
% \end{theorem}

% \begin{theorem}[$\llangle - \rrangle$ is faithful]
% \end{theorem}

% \update{Or we can define rewriting for string diagrams in $\textsf{PROP}^{+}(\Sigma)$ based on rewriting in $\textsf{PROP}(\Sigma)$ and then say that $f \xrightarrow{}_{\mathcal{R}} g$ iff $[f] [g]$}

@@

\begin{theorem}
    $a \Rightarrow_{\langle l, r \rangle}^{+} b$ in $\textsf{PROP}^{+}$ iff $[a] \Rrightarrow_{\langle [l],[r] \rangle} [b]$ in $\sfrac{\WellTypedMdaEcospans}{\mathcal{S}}$.
\end{theorem}
